DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-07-21 2015CompleteDerivation/completeDerivation.tex
Added status and reminder to download latest version
Fixed symplectic manifold justification and other edits

\begin{abstract}
\textbf{This manuscript is a work in progress.} Ideas are constantly reshaped to find more precise and elegant arguments. It is provided as is to stimulate discussion.

@@

\section{Introduction}

TODO: Point out most modern physics starts by assuming some mathematical framework. Pros and cons of that.

TODO: High level view of the derivation

@@

\begin{prop}\label{differentiable_manifold}
	The state space $\mathcal{S}$ for the particles of a  classical material identified by a finite number of continuous independent state variables is a differentiable manifold. The state space $\mathcal{C}$ for a homogeneous classical material is isomorphic to the space of Lebesgue integrable differentiable functions. That is $\mathcal{C} \cong C^1(\mathcal{S}) \cap L^1(\mathcal{S}, \mu)$, where $C^1(\mathcal{S}) \equiv \{\rho:\mathcal{S} \rightarrow \mathbbm{R} \; | \; \rho$ is differentiable$\}$.
\end{prop}
\begin{justification}
	We claim $\mathcal{S}$ is a differential manifold. Let $\rho_\mathcal{c} : \mathcal{S} \rightarrow \mathbbm{R}$ be the distribution associated with a composite state $\mathcal{c} \in \mathcal{C}$. $\rho_\mathcal{c}$ is defined on any chart of $\mathcal{S}$. Given two charts, for any $\rho_\mathcal{c}(q)$ there is a corresponding $\rho_\mathcal{c}(q')=\rho_\mathcal{c}(q^i) | d q' / d q |$. The Jacobian $| d q' / d q |$ exists and is non-zero. The map between any two charts is differentiable. $\mathcal{S}$ is a differentiable manifold.

@@

Note that neither of those expressions match the norm given by the vector product defined as $\langle f, g \rangle = \int_{\mathcal{S}} fg \, d\mu$. Moreover, since $\mathcal{C}$ excludes discontinuous functions, it can never be a complete normed space: it will not include the limit for all Cauchy sequences. Such constructions, though, are still useful. Consider the following expressions:
\begin{align*}
\Lambda_{\mathcal{S}} (\mathcal{c}) &= \int_\mathcal{S} \rho_{\mathcal{c}} d \mu = \int_\mathcal{S} 1\cdot  \rho_{\mathcal{c}} d \mu \\
\Lambda_U (\mathcal{c}) &= \int_U \rho_{\mathcal{c}} d \mu = \int_\mathcal{S} 1_U \cdot \rho_{\mathcal{c}} d \mu \\
p_{total}(\mathcal{c}) &= \int_\mathcal{S} p \cdot \rho_{\mathcal{c}} d \mu \\
q_{avg}(\mathcal{c}) &= 1/|\rho_{\mathcal{c}}| \int_\mathcal{S} q \cdot \rho_{\mathcal{c}} d \mu
\end{align*}
For each state, they represent respectively the total amount of material, the material within $U$, the total momentum and the average of a state variable. As they are linear functionals of $\mathcal{C}$,  they can be expressed using their dual vector over $\mathcal{S}$. But these vectors ($1$, $1_U$, $p$, $q/|\rho_{\mathcal{c}}|$) are \emph{not} elements of $\mathcal{C}$. They do not represent states, they are not continuous and integrable, and there is no single general physical meaning for them. Therefore, as we mentioned before, while we can extend the function space for convenience, computation purposes or to study a limit case, we need to be mindful of the extension and carefully check that the mathematical results are physically meaningful case by case.

@@

The simplest case is when a single variable is enough to define our units, and therefore tell how all the others must change. Suppose that our state space $\mathcal{S}$ is identified by $n$ independent state variables $\{q, k^i\}$ and a change of units for the first, that is $\hat{q} = \hat{q}(q)$, determines the change for all others. The Jacobian matrix is the block matrix:
\begin{align*}
J =  \left[
\begin{array}{cc}
d_q\hat{q} & 0 \\
\partial_q \hat{k}^j & \partial_i \hat{k}^j \\
\end{array}
\right] 
\end{align*}
The Jacobian has to be unitary, therefore all elements $\partial_i \hat{k}^j$ must be well defined and $|\partial_i \hat{k}^j| = 1/d\hat{q}$. This only happens if $\partial_i \hat{k}^j$ is a $1\times1$ matrix: there is only one $k$. This gives means $\partial_{k} \hat{k} = d_{\hat{q}} q$ and $\hat{k} = d_{\hat{q}} q \, k + f(q)$. As this is a change of units, we expect the zero value of $k$ to remain constant: $\hat{k}(q, 0) = 0 = f(q)$. Therefore $\hat{k} = d_{\hat{q}} q \, k$ transforms as the component of a covariant vector. $\mathcal{S}$ is isomorphic to $T^*\mathcal{Q}$ where $\mathcal{Q}$ is the manifold that defines our unit.

@@

\begin{justification}
	We claim the simplest state space $\mathcal{S}$ that allows state-variable-invariant densities is the symplectic manifold $(T^*\mathcal{Q}, \omega)$ where $T^*\mathcal{Q}$ is the cotangent bundle of a one dimensional differential manifold $\mathcal{Q}$ and $\omega$ is the canonical two-form. The description of the simplest physical object on which we can define a state-variable-invariant must be given by at least one continuous state variable $q$. Let $\mathcal{Q}$ be the manifold charted by $q$. $\mathcal{Q}$ is an embedded manifold of $\mathcal{S}$. $\mathcal{Q}$ is differentiable. Let $q$ fully define the unit system for the description of the physical object. Then the physical object is represented by a geometrical object on $\mathcal{Q}$. The physical object is infinitesimal, therefore is represented by a local geometrical object. The simplest geometrical object in $\mathcal{Q}$ is a point, charted by $q$. A density over points is not invariant: $\rho(q')_\mathcal{c}= \rho(q)_\mathcal{c} dq/dq'$. Next we consider local geometrical objects charted by 2 variables. Densities over vectors $v(q) e_q \in T\mathcal{Q}$ are not invariant: $\rho_\mathcal{c}(q', v')= \rho(q, v)_\mathcal{c} (dq/dq')^2$. Densities over covectors $k(q) dq \in T^*\mathcal{Q}$ are invariant: $\rho_\mathcal{c}(q', k')= \rho_\mathcal{c}(q, k) (dq/dq') (dq'/dq) = \rho_\mathcal{c}(q, k)$. Covectors are the simplest geometrical object to allow unit independent densities. The state space $\mathcal{S}$ for the simplest object that allows state-variable-independent densities is $T^*\mathcal{Q}$ of a one dimensional differential manifold $\mathcal{Q}$. Express integration in coordinates. $\int_{\mathcal{S}} \rho_\mathcal{c} d\mu \propto \int_{\mathcal{S}} \rho_\mathcal{c}(q, k) dq \wedge dk$. $d\mu = \hbar dq \wedge dk = dq \wedge dp = \omega$ where $\hbar$ is a constant, $p\equiv \hbar k$  and $\omega$ is the canonical two form. $\omega$ is invariant under state variable change. $\omega$ is a symplectic metric for $\mathcal{S}=T^*\mathcal{Q}$. $\mathcal{S} = (T^*\mathcal{Q}, \omega)$ is a symplectic manifold.

@@

We recognize the familiar set of Hamilton's equations. They are the set of equations to describe the deterministic and reversible motion of the infinitesimal parts of an infinitesimally reducible homogeneous material. In other words: the forces that conserve energy (i.e. the value of a suitable Hamiltonian) are exactly the ones that provide deterministic and reversible motion. The challenge in their derivation mainly lies in the necessary use of different branches of mathematics (e.g. topology, measure theory, differential geometry, symplectic geometry). The conceptual meaning, on the other hand, is hopefully straight forward.

>>>>

\begin{abstract}
\textbf{This manuscript is a work in progress.} Ideas are constantly reshaped to find more precise and elegant arguments. It is provided as is to stimulate discussion.  \textbf{Make sure you have the latest version from http://assumptionsofphysics.org}

@@

\section{Status}

Section III - Done.

Section IV - Done.

Section V - Mostly done. Needs review.

Section VI - Rough draft

%TODO: Point out most modern physics starts by assuming some mathematical framework. Pros and cons of that.

%TODO: High level view of the derivation

@@

\begin{prop}\label{differentiable_manifold}
	The state space $\mathcal{S}$ for the particles of a  classical material identified by a finite number of continuous independent state variables is a differentiable manifold. The state space $\mathcal{C}$ for a homogeneous classical material is isomorphic to the space of Lebesgue integrable differentiable functions. That is $\mathcal{C} \cong C^1(\mathcal{S}) \cap L^1(\mathcal{S}, \mu)$, where $C^1(\mathcal{S}) \equiv \{\rho:\mathcal{S} \rightarrow \mathbbm{R} \; | \; \rho$ is differentiable$\}$.
\end{prop}
\begin{justification}
	We claim $\mathcal{S}$ is a differential manifold. Let $\rho_\mathcal{c} : \mathcal{S} \rightarrow \mathbbm{R}$ be the distribution associated with a composite state $\mathcal{c} \in \mathcal{C}$. Let $q^i$ and $\hat{q}^j$ two sets of coordinates. Let $\rho_\mathcal{c}(q^i)$ and $\rho_\mathcal{c}(\hat{q}^j)$ be the expression in local coordinates of the distribution.  $\rho_\mathcal{c}(\hat{q}^j)=\rho_\mathcal{c}(q^i) | \partial_j q^i |$ as it transforms as a density. $\rho_\mathcal{c}(\mathcal{s}) = \rho_\mathcal{c}(q^i(\mathcal{s})) = \rho_\mathcal{c}(\hat{q}^j(\mathcal{s}))$. The Jacobian $| d q' / d q |$ exists and is non-zero. The map between any two charts is differentiable. $\mathcal{S}$ is a differentiable manifold.

@@

Note that neither of those expressions match the norm given by the vector product defined as $\langle f, g \rangle = \int_{\mathcal{S}} fg \, d\mu$. Moreover, since $\mathcal{C}$ excludes discontinuous functions, it can never be a complete normed space: it will not include the limit for all Cauchy sequences, it cannot be a Banach or a Hilbert space. Such constructions, though, are still useful. Consider the following expressions:
\begin{align*}
\Lambda_{\mathcal{S}} (\mathcal{c}) &= \int_\mathcal{S} \rho_{\mathcal{c}} d \mu = \langle 1 , \rho_{\mathcal{c}} \rangle \\
\Lambda_U (\mathcal{c}) &= \int_U \rho_{\mathcal{c}} d \mu = \langle 1_U , \rho_{\mathcal{c}} \rangle \\
\rho_\mathcal{c}(q_0,p_0) &= \int_\mathcal{S} \delta_{q_0,p_0} \rho_{\mathcal{c}} d \mu = \langle \delta_{q_0,p_0} , \rho_{\mathcal{c}} \rangle \\
p_{total}(\mathcal{c}) &= \int_\mathcal{S} p \cdot \rho_{\mathcal{c}} d \mu = \langle p , \rho_{\mathcal{c}} \rangle \\
q_{avg}(\mathcal{c}) &= \frac{\int_\mathcal{S} q \cdot \rho_{\mathcal{c}} d \mu}{\int_\mathcal{S} \rho_{\mathcal{c}} d \mu} = \frac{\langle q , \rho_{\mathcal{c}} \rangle}{\langle 1 , \rho_{\mathcal{c}} \rangle}
\end{align*}
For each state, they represent respectively the total amount of material, the material within $U$, the density at $(q_0, p_0)$, the total momentum and the average of a state variable. As they are linear functionals of $\mathcal{C}$,  they can be expressed using their dual vector over $\mathcal{S}$. But these vectors ($1_U$, $p$, ...) are \emph{not} elements of $\mathcal{C}$. They do not represent states, they are not continuous and integrable, and there is no single general physical meaning for all of them. Therefore, as we mentioned before, while we can extend the function space for convenience, computation purposes or to study a limit case, we need to be mindful of the extension and carefully check that the mathematical results are physically meaningful case by case.

@@

The simplest case is when a single variable is enough to define our units, and therefore tell how all the others must change. Suppose that our state space $\mathcal{S}$ is identified by $n$ independent state variables $\{q, k^i\}$ and a change of units for the first, that is $\hat{q} = \hat{q}(q)$, determines the change for all others. The Jacobian matrix is the block matrix:
\begin{align*}
J =  \left[
\begin{array}{cc}
d_q\hat{q} & 0 \\
\partial_q \hat{k}^j & \partial_i \hat{k}^j \\
\end{array}
\right] 
\end{align*}
The Jacobian has to be unitary, therefore all elements $\partial_i \hat{k}^j$ must be well defined and $|\partial_i \hat{k}^j| = 1/d\hat{q}$. This only happens if $\partial_i \hat{k}^j$ is a $1\times1$ matrix: there is only one $k$. This gives $\partial_{k} \hat{k} = d_{\hat{q}} q$ and $\hat{k} = d_{\hat{q}} q \, k + f(q)$. As this is a change of units, we expect the zero value of $k$ to remain constant: $\hat{k}(q, 0) = 0 = f(q)$. Therefore $\hat{k} = d_{\hat{q}} q \, k$ transforms as the component of a covariant vector. $\mathcal{S}$ is isomorphic to $T^*\mathcal{Q}$ where $\mathcal{Q}$ is the manifold that defines our unit.

@@

\begin{justification}
	We claim the simplest state space $\mathcal{S}$ that allows state-variable-invariant densities is the cotangent bundle $T^*\mathcal{Q}$ of a single dimensional manifold $\mathcal{Q}$. Let $\mathcal{S}$ be the $n$ dimensional manifold for the particle states. $\mathcal{S}$ must be at least one dimensional to allow state-variable-invariant densities. Consider a set of state variables for $U \in \mathcal{S}$: the unit system for the simplest state space will be fully described by a single state variable. Let $q$ be the state variable that defines the unit system. Let $\mathcal{Q}$ be the manifold charted by $q$. $\mathcal{Q}$ is an embedded manifold of $\mathcal{S}$. $\mathcal{Q}$ is differentiable. A diffeomorphism $\hat{q}=\hat{q}(q)$ fully defines a unit transformation. Let $(q,k^i)$ with $i=1...n-1$ be a set of state variable. The Jacobian determinant of the transformation is $|J| = d_q\hat{q} |\partial_i\hat{k}^j|$. $|J|=1$ as densities must be invariant. $|\partial_i \hat{k}^j| = d_{\hat{q}}q$. $\partial_i \hat{k}^j$ is one dimensional as its components must be fully specified by the previous relationship. Let $k=k^1$. $\hat{k}(q,0) = 0$ as a change in units does not change the $0$ value. $\partial_k \hat{k} = d_{\hat{q}}q$. $\hat{k} = d_{\hat{q}}q k$. $k$ changes as the component of a co-vector. Each state is identified by a point and a co-vector in $\mathcal{Q}$. The simplest state space $\mathcal{S}$ is isomorphic to $T^*\mathcal{Q}$ where $Q$ is the one dimensional manifold that defines the units.

	We claim the simplest state space $\mathcal{S}$ that allows state-variable-invariant densities is the symplectic manifold $(T^*\mathcal{Q}, \omega)$ where $T^*\mathcal{Q}$ is the cotangent bundle of a one dimensional differential manifold $\mathcal{Q}$ and $\omega$ is the canonical two-form. Express integration in coordinates. $\int_{\mathcal{S}} \rho_\mathcal{c} d\mu \propto \int_{\mathcal{S}} \rho_\mathcal{c}(q, k) dq \wedge dk$. $d\mu = \hbar dq \wedge dk = dq \wedge dp = \omega$ where $\hbar$ is a constant, $p\equiv \hbar k$  and $\omega$ is the canonical two form. $\omega$ is invariant under state variable change. $\omega$ is a symplectic metric for $\mathcal{S}=T^*\mathcal{Q}$. $\mathcal{S} = (T^*\mathcal{Q}, \omega)$ is a symplectic manifold.

@@

We recognize the familiar set of Hamilton's equations. They are the set of equations that describe the deterministic and reversible motion of the infinitesimal parts of an infinitesimally reducible homogeneous material. In other words: the forces that conserve energy (i.e. the value of a suitable Hamiltonian) are exactly the ones that provide deterministic and reversible motion. The challenge in their derivation mainly lies in the necessary use of different branches of mathematics (e.g. topology, measure theory, differential geometry, symplectic geometry). The conceptual meaning, on the other hand, is hopefully straight forward.
